Simple repression draft

### 2020-05-26 doc/abstract.tex
RP work on May 26
I worked on the Abstract, the Intro, the Discussion and put in a label in Section 4 so when it was called the section got the right number.

% !TEX root = ./busty_transcription.tex
\begin{abstract}
The study of transcription remains one of the centerpieces of modern biology
with implications in settings from development to metabolism to evolution to
disease. Precision measurements using a host of different techniques including
fluorescence and sequencing readouts have raised the bar for what it means to
quantitatively understand transcriptional regulation. In particular our
understanding of the simplest genetic circuit is sufficiently refined both
experimentally and theoretically that it has become possible to carefully
discriminate between different conceptual pictures of how this regulatory system
works. This regulatory motif, originally posited by Jacob and Monod in the
1960s, consists of a single transcriptional repressor binding to a promoter site
and inhibiting transcription. In this paper, we show how seven distinct models
of this so-called simple-repression motif, based both on equilibrium and kinetic
thinking, can be used to derive the predicted levels of gene expression. These
different models are invoked to confront a variety of different data on mean,
variance and full gene expression distribution, illustrating the extent to which
such models can be distinguished.
\end{abstract}

>>>>

% !TEX root = ./busty_transcription.tex
\begin{abstract}
The study of transcription remains one of the centerpieces of modern biology
with implications in settings from development to metabolism to evolution to
disease. Precision measurements using a host of different techniques including
fluorescence and sequencing readouts have raised the bar for what it means to
quantitatively understand transcriptional regulation. In particular our
understanding of the simplest genetic circuit is sufficiently refined both
experimentally and theoretically that it has become possible to carefully
discriminate between different conceptual pictures of how this regulatory system
works. This regulatory motif, originally posited by Jacob and Monod in the
1960s, consists of a single transcriptional repressor binding to a promoter site
and inhibiting transcription. In this paper, we show how seven distinct models
of this so-called simple-repression motif, based both on equilibrium and kinetic
thinking, can be used to derive the predicted levels of gene expression. These
different models are invoked to confront a variety of different data on mean,
variance and full gene expression distribution, illustrating the extent to which
such models can be distinguished, and suggesting a two-state model with a distribution
of burst sizes as the most potent of the seven for describing the simple-repression motif.
\end{abstract}

### 2020-05-26 doc/section_01_introduction.tex
RP work on May 26
I worked on the Abstract, the Intro, the Discussion and put in a label in Section 4 so when it was called the section got the right number.

To that end, in this paper we examine a wide variety of distinct models for the
simple repression regulatory architecture. This genetic architecture consists of
a DNA promoter regulated by a transcriptional repressor that binds to a single
binding site \cite{Garcia2011a}. All of the proposed models models coarse-grain
away some of the important microscopic features of this architecture that have
been elucidated by generations of geneticists, molecular biologists and
biochemists.
\mrm{Our aim is to understand the underlying principles of such coarse-graining
in order to increase the predictive power of these models. In doing so we hope
future models of regulatory response will be able to serve the powerful
predictive role needed to take synthetic biology from a brilliant exercise in
enlightened empiricism to a rational design framework as any other branch of
engineering.}
% Our aim is to understand the underlying principles of such
% coarse-graining such that future models of regulatory response can serve the
% powerful predictive role needed to take synthetic biology from what is often a
% brilliant exercise in enlightened empiricism and recast it as the same kind of
% rational design process we are used to in other branches of engineering. 
More precisely, we want phenomenology in the sense of coarse-graining away
atomistic detail, but still retaining biophysical meaning. For example, we are
not satisfied with the strictly phenomenological approach offered by the
commonly used Hill functions. Studies like~\cite{Razo-Mejia2018} have
demonstrated that Hill functions are clearly insufficient since each new
situation requires a completely new set of parameters. Such work requires a
quantitative theory of how biophysical changes at the molecular level propagate
to input-output functions at the genetic circuit level. We want concepts, not
mere facts. In particular a key question is: at this level of coarse graining,
what microscopic details do we need to explicitly model, and how do we figure
that out? For example, do we need to worry about all or even any of the steps
that individual RNAPs go through each time they make a transcript? Turning the
question around, can we see any imprint of those processes in the available
data? If the answer is no, then those processes are irrelevant for our purposes.
Forward modeling and inverse (statistical inferential) modeling are necessary to 
tackles such questions.

@@

The logic of the remainder of the paper is as follows. In
section~\ref{section_02_means}, we show how both thermodynamic models and
kinetic models based upon the chemical master equation all culminate in the same
underlying functional form for the fold-change in the average level of gene
expression as shown in Figure~\ref{fig1:means_cartoons}(D).
Section~\ref{section_03_beyond_means.tex} goes beyond an analysis of the mean
gene expression by asking how the same models presented in
Figure~\ref{fig1:means_cartoons}(C) can be used to explore noise in gene
expression. To make contact with experiment, all of these models must make a
commitment to some numerical values for the key parameters found in each such
model. Therefore in Section~\ref{section_04_bayesian_inference.tex} we explore
the use of Bayesian inference to establish these parameters and to rigorously
answer the question of how to discriminate between the different models.

>>>>

To that end, in this paper we examine a wide variety of distinct models for the
simple repression regulatory architecture. This genetic architecture consists of
a DNA promoter regulated by a transcriptional repressor that binds to a single
binding site \cite{Garcia2011a}. All of the proposed models models coarse-grain
away some of the important microscopic features of this architecture that have
been elucidated by generations of geneticists, molecular biologists and
biochemists.
Our aim is to understand the underlying principles of such coarse-graining
in order to increase the predictive power of these models. In doing so we hope
future models of regulatory response will be able to serve the powerful
predictive role needed to take synthetic biology from a brilliant exercise in
enlightened empiricism to a rational design framework as  in any other branch of
engineering.
% Our aim is to understand the underlying principles of such
% coarse-graining such that future models of regulatory response can serve the
% powerful predictive role needed to take synthetic biology from what is often a
% brilliant exercise in enlightened empiricism and recast it as the same kind of
% rational design process we are used to in other branches of engineering. 
More precisely, we want phenomenology in the sense of coarse-graining away
atomistic detail, but still retaining biophysical meaning. For example, we are
not satisfied with the strictly phenomenological approach offered by the
commonly used Hill functions. Studies like~\cite{Razo-Mejia2018} have
demonstrated that Hill functions are clearly insufficient since each new
situation requires a completely new set of parameters. Such work requires a
quantitative theory of how biophysical changes at the molecular level propagate
to input-output functions at the genetic circuit level. We want concepts, not
mere facts. In particular a key question is: at this level of coarse graining,
what microscopic details do we need to explicitly model, and how do we figure
that out? For example, do we need to worry about all or even any of the steps
that individual RNAPs go through each time they make a transcript? Turning the
question around, can we see any imprint of those processes in the available
data? If the answer is no, then those processes are irrelevant for our purposes.
Forward modeling and inverse (statistical inferential) modeling are necessary to 
tackles such questions.

@@

The logic of the remainder of the paper is as follows. In
section~\ref{section_02_means}, we show how both thermodynamic models and
kinetic models based upon the chemical master equation all culminate in the same
underlying functional form for the fold-change in the average level of gene
expression as shown in Figure~\ref{fig1:means_cartoons}(D).
Section~\ref{sec:beyond_means} goes beyond an analysis of the mean
gene expression by asking how the same models presented in
Figure~\ref{fig1:means_cartoons}(C) can be used to explore noise in gene
expression. To make contact with experiment, all of these models must make a
commitment to some numerical values for the key parameters found in each such
model. Therefore in Section~\ref{section_04_bayesian_inference} we explore
the use of Bayesian inference to establish these parameters and to rigorously
answer the question of how to discriminate between the different models.

### 2020-05-26 doc/section_04_bayesian_inference.tex
RP work on May 26
I worked on the Abstract, the Intro, the Discussion and put in a label in Section 4 so when it was called the section got the right number.

% !TEX root = ./bursty_transcription.tex
\section{Finding the ``right'' model: Bayesian parameter inference and model 
selection}
\subsection{Parameter inference for constitutive promoters}

@@

\mmnote{Outline:
\begin{itemize}
\item Hey look it mostly works. With full model, 9D model is well identified,
though most of the individual experiments unsurprisingly are not:
only 3 of the 9 datasets were identifiable if fit in isolation
and for the rest only the ratio of repressor rates is identifiable.
Only by making assumptions that certain rates
are equal across datasets was identifiability achieved.
PPC is not perfect, it kinda misses some
rep/op pairs but overall it's surprisingly predictive considering how strong
were the assumptions that went into it.
\item Do we chalk this up to experimental imperfections or is there interesting
biology to be found in the disagreements? I'm inclined towards the former. Even
just catching cells in truly apples-to-apples growth phases, or matching up
parameters of imaging sessions weeks or months apart, is hardly trivial.
\item The alternative is to refine our model and/or add on complexity, and it's
not at all clear what additions to make to the model, especially how to produce
the long tail of repressed strains over UV5. I see no way to get that without
getting into complicated time-correlation/history effects, where
transcription...rebounds?? After being repressed?? Sounds odd.
\item Comparison between my inferred rates, equilibrium binding E, and
single-molecule measurements. Not quite as slam-dunk as I'd hoped because of
$\gamma$ dependence, but rates are w/in factor of 2 or better, seems quite good
really! With exception of one Oid data data point, free energies are w/in about
0.5 kT also. That's not massively larger than the repeatability threshold of
fold-change measurements, at maybe 0.1 to 0.3 kT.
\item Surprising that so many model parameters can be inferred from just a few
distributions that, by eye, don't have strong features! Does require some
assumptions, e.g., about equivalence of rates across experiments, but that seems
quite reasonable.
\end{itemize}
}

>>>>

% !TEX root = ./bursty_transcription.tex
\section{Finding the ``right'' model: Bayesian parameter inference and model 
selection}
\label{section_04_bayesian_inference}
\subsection{Parameter inference for constitutive promoters}

@@

### 2020-05-26 doc/section_05_discussion.tex
RP work on May 26
I worked on the Abstract, the Intro, the Discussion and put in a label in Section 4 so when it was called the section got the right number.

Our key insights can be summarized as follows. First, as shown in 
Figure~\ref{fig1:means_cartoons}, the mean expression in the simple repression
architecture is captured by a master curve in which the action of repressor and
the details of the polymerase interaction with the promoter appear separately
and additively in an effective free energy. Interestingly, as has been shown
elsewhere in the context of the Monod-Wyman-Changeux model, these kinds of
coarse-graining results are an exact mathematical result and do not constitute
hopeful approximations or biological naivete \cite{Razo-Mejia2018, Chure2019}.
To further dissect the relative merits of the different models, we must appeal
to higher moments of the gene expression probability distribution. To that end,
our second set of insights focus on gene expression noise, where it is seen that
a treatment of the constitutive promoter already reveals that some models have
Fano factors (standard deviation/mean) that are less than one, at odds with
any and all experimental data that we are aware of.
\mrm{
This theoretical result allows us to directly discard a subset of the models
(models 1-3 in Figure~\ref{fig2:fig2:constit_cartoons}) since they cannot be
reconciled with experimental observations. The two models left (models 4 and 5
in Figure~\ref{fig2:fig2:constit_cartoons}) seem to provide a rich enough
phenomenology to be able to reproduce the data. A previous exploration of one 
of these models demonstrated the sloppy nature of the model in which single-cell
mRNA counts data alone cannot constrain the value of all parameters 
simultaneously \cite{Razo-Mejia2020}. Here we demonstrate that the proposed
one-state bursty promoter model (model 5 in 
Figure~\ref{fig2:fig2:constit_cartoons}) is a limit of the commonly used
two-state promoter model \cite{Peccoud1995}. We put at test the idea that this
level of coarse-graining is descriptive enough to reproduce previous 
experimental observations. In particular we perform Bayesian inference on the
two parameters describing the full steady-state mRNA distribution, finding that
the model is able to capture a plethora of promoter sequences with different
mean levels of expression and noise.
}

\mrm{
Here we should include a discussion of the regulated case.
}

@@

>>>>

Our key insights can be summarized as follows. First, as shown in 
Figure~\ref{fig1:means_cartoons}, the mean expression in the simple repression
architecture is captured by a master curve in which the action of repressor and
the details of the polymerase interaction with the promoter appear separately
and additively in an effective free energy. Interestingly, as has been shown
elsewhere in the context of the Monod-Wyman-Changeux model, these kinds of
coarse-graining results are an exact mathematical result and do not constitute
hopeful approximations or biological naivete \cite{Razo-Mejia2018, Chure2019}.
To further dissect the relative merits of the different models, we must appeal
to higher moments of the gene expression probability distribution. To that end,
our second set of insights focus on gene expression noise, where it is seen that
a treatment of the constitutive promoter already reveals that some models have
Fano factors (standard deviation/mean) that are less than one, at odds with
any and all experimental data that we are aware of.
This theoretical result allows us to directly discard a subset of the models
(models 1-3 in Figure~\ref{fig2:constit_cartoons}) since they cannot be
reconciled with experimental observations. The two remaining models (models 4 and 5
in Figure~\ref{fig2:constit_cartoons}) appear to contain enough microscopic
realistm to be able to reproduce the data. A previous exploration of one 
of these models demonstrated the sloppy nature of the model in which data on single-cell
mRNA counts alone cannot constrain the value of all parameters 
simultaneously \cite{Razo-Mejia2020}. Here we demonstrate that the proposed
one-state bursty promoter model (model 5 in 
Figure~\ref{fig2:constit_cartoons}) emerges as  a limit of the commonly used
two-state promoter model \cite{Peccoud1995}. We put  the idea to the test that this
level of coarse-graining is rich enough to reproduce previous 
experimental observations. In particular we perform Bayesian inference to determine the
two parameters describing the full steady-state mRNA distribution, finding that
the model is able to provide a quantitative description of  a plethora of promoter sequences with different
mean levels of expression and noise.

\mrm{
Here we should include a discussion of the regulated case.  RP: I made a first
brief stab at it.
}
With the results of the constitutive promoter in hand, we then fix the parameters
associated with this class of promoters and use them as input for evaluating
the noise in gene expression for the simple repression motif itself.  This analysis
allows us to provide a single overarching analysis of both the constitutive
and simple repression architectures using one simple model and
corresponding set of self-consistent parameters, demonstrating not only
a predictive framework, but also reconciling the equilibrium and non-equilibrium
views of the same simple repression constructs.

@@

\mmnote{Outline:
\begin{itemize}
\item Hey look it mostly works. With full model, 9D model is well identified,
though most of the individual experiments unsurprisingly are not:
only 3 of the 9 datasets were identifiable if fit in isolation
and for the rest only the ratio of repressor rates is identifiable.
Only by making assumptions that certain rates
are equal across datasets was identifiability achieved.
PPC is not perfect, it kinda misses some
rep/op pairs but overall it's surprisingly predictive considering how strong
were the assumptions that went into it.
\item Do we chalk this up to experimental imperfections or is there interesting
biology to be found in the disagreements? I'm inclined towards the former. Even
just catching cells in truly apples-to-apples growth phases, or matching up
parameters of imaging sessions weeks or months apart, is hardly trivial.
\item The alternative is to refine our model and/or add on complexity, and it's
not at all clear what additions to make to the model, especially how to produce
the long tail of repressed strains over UV5. I see no way to get that without
getting into complicated time-correlation/history effects, where
transcription...rebounds?? After being repressed?? Sounds odd.
\item Comparison between my inferred rates, equilibrium binding E, and
single-molecule measurements. Not quite as slam-dunk as I'd hoped because of
$\gamma$ dependence, but rates are w/in factor of 2 or better, seems quite good
really! With exception of one Oid data data point, free energies are w/in about
0.5 kT also. That's not massively larger than the repeatability threshold of
fold-change measurements, at maybe 0.1 to 0.3 kT.
\item Surprising that so many model parameters can be inferred from just a few
distributions that, by eye, don't have strong features! Does require some
assumptions, e.g., about equivalence of rates across experiments, but that seems
quite reasonable.
\end{itemize}
}
